---
title: 'Modeling Lake Trophic State: A Data Mining Approach'
author:
- affiliation: aed
  email: hollister.jeff(at)epa.gov
  footnote: Corresponding author
  name: Jeffrey W. Hollister
- affiliation: aed
  name: W. Bryan Milstead
- affiliation: aed
  name: Betty J. Kreakie
address:
- address: "US Environmental Protection Agency Office of Research and Development National
    Health and Environmental Effects Research Laboratory Atlantic Ecology Division 27
    Tarzwell Drive  Narragansett, RI, 02879, USA  "
  code: aed
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: elsarticle.latex
  html_document: null
  word_document: null
csl: plos.csl
documentclass: elsarticle
layout: review
linenumbers: no
bibliography: LakeTrophicModelling.bib
abstract:  Productivity of lentic ecosystems has been well studied and it is 
  widely accepted that as nutrient inputs increase,  productivity increases and
  lakes transition from low trophic state (e.g. oligotrophic) to higher trophic 
  states  (e.g. eutrophic). These broad trophic state classifications are good 
  predictors of ecosystem health and ecosystem services/disservices (e.g. recreation,
  aesthetics, fisheries, and harmful algal blooms). While the relationship between 
  nutrients and trophic state provides reliable predictions, it requires *in situ* 
  water quality data in order to paramterize the model.  This limits the application 
  of these models to lakes with existing and, more importanly, available water quality
  data.  To expand our ability to predict in lakes without water quality data, we take
  advantage of the availability of a large national lakes water quality database, 
  land use/land cover data, lake morphometry data, other universally available data, 
  and modern data mining approaches to build and assess models of lake tropic state 
  that may be more universally applied.  We use random forests and random forest 
  variable selection to identify variables to be used for predicting trophic state and 
  we compare the performance of two models of trophic state (as determined by 
  chlorophyll a concentration).   The first model estimates trophic state with *in situ* 
  as well as universally available data and the second model uses universally available 
  data only. For each of these models we used three separate trophic state categories, 
  for a total of six models. Overall accuracy for the *in situ* and universal data 
  models ranged from xx% to xx% and xx, xx, and xx described the most variation in 
  trophic state.  For the universal data only models, Overall accuraccy ranged from xx% 
  to xx% and xx, xx, and xx described the most variation in trophic state. Lastly, it 
  is believed that the presence and abundance of cyanobacteria is strongly associated 
  with trophic state.  To test this we examine the association between estimates of 
  cyanobacteria biovolume and the measured and predicted trophic state.  Expanding 
  these preliminary results to include cyanobacteria taxa indicates that cyanobacteria 
  are significantly more likely to be found in highly eutrophic lakes. These results 
  suggest that predictive models of lake trophic state may be improved with additional 
  information on the landscape surrounding lakes and that those models provide 
  additional information on the presence of potentially harmful cyanobacteria taxa. 
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Modeling Lake Trophic State: A Data Mining Approach}
-->

```{r setup, include=FALSE, echo=FALSE}
devtools::install_github("LakeTrophicModelling","USEPA",quick=TRUE,
                         auth_token="8742a8c1c2207393659a1de1bd53527fa56ea0d7")
devtools::install_github("wesanderson","jhollist")
library("wesanderson")
library("LakeTrophicModelling")
library("knitr")
library("ggplot2")
library("sp")
library("rgdal")
library("xtable")
library("e1071")
#opts_chunk$set(dev = 'pdf', fig.width=6, fig.height=4.3)
data(hkm2014Data_all)
#Checks for existing cache (from another project)
if(file.exists("cache")){
  for(i in gsub(".rdb","",list.files("cache",".rdb",full.names=T))){
    lazyLoad(i)
  }
}
```

#Introduction
Cyanobacteria are an important taxonomic group associated with harmful algal blooms in lakes. Understanding the drivers of cyanobacteria presence has important implications for lake management and for
 the protection of human and ecosystem health.  Chlorophyll a concentration, a measure of the biological productivity of a lake, is one such driver and is largely, although not exclusively, determined
 by nutrient inputs.   As nutrient inputs increase, productivity increases and lakes transition from low trophic state (e.g. oligotrophic) to higher trophic states (e.g. hypereutrophic).  These broad
trophic state classifications are associated with ecosystem health and ecosystem services/disservices (e.g. recreation, aesthetics, fisheries, and harmful algal blooms).  Thus, models of trophic state
 might be used to predict things like cyanobacteria.

We have three goals for this preliminary research:

1. Build and assess models of lake trophic state
2. Assess ability to predict trophic state in lakes without available *in situ* water quality data
3. Explore association between cyanobacteria and trophic in order to expand models.

#Data and Modeling Methods

##Data

We utilize four primary sources of data for this study. These are outlined below and in Table 1.

1. National Lakes Assessment (NLA) 2007: The NLA data were collected during the summer of 2007 and the final data were released in 2009.  With consistent methods and metrics collected at 1056 location
s across the conterminous United States (Map 1), the NLA provides a unique opportunity to examine broad scale patterns in lake productivity.  The NLA collected data on biophysical measures of lake wat
er quality and habitat.  For this analysis we primarily examined the water quality measurements from the NLA [@usepa2009national].
2. National Land Cover Dataset (NLCD) 2006: The NLCD is a nationally collected land use land cover dataset.  We collected total land use land cover and total percent impervious surface within a 3 kilo
meter buffer surrounding the lake to examine larger landscape-level effects [@homer2004development;@xian2009updating].
3. Modeled lake morphometry: Various measures of lake morphometry (i.e. depth, volume, fetch, etc.) are important in understanding lake productivity, yet many of these data are difficult to obtain for
 large numbers of lakes over broad regions.  To add this information we modeled lake morphometry [@hollister2010volume;@hollister_predicting_2011;@Lakmorpho2014;@NLMDInPrep].
4. Estimated Cyanobacteria Biovolumes: Cyanobacteria biovolumes are a truer measure of Cyanobacteria dominance than abundance as there is great variability in the size within and between species.  To a
ccount for this, Beaulieu *et al.* [-@beaulieu2013nutrients] used literature values to estimate biovolumes for the taxa in the NLA.  They shared this data and we have summed that information on a per-
lake basis.

```{r Table1, results='asis', echo=FALSE}
translation<-c("Percent Impervious","Percent Water","Percent Ice/Snow","Percent Developed Open Space",
               "Percent Low Intensity Development","Percent Medium Intensity Development",
               "Percent High Intensity Development","Percent Barren","Percent Decidous Forest",
               "Percent Evergreen Forest","Percent Mixed Forest","Precent Shrub/Scrub","Percent Grassland",
               "Percent Pasture", "Percent Cropland", "Percent Woody Wetland", 
               "Percent Herbaceuos Wetland","Longitude","Latitude","Lake Surface Area", "Lake Perimeter",
               "Shoreline Development Index","Date Samples Collected", "Ecoregion", "Watershed Area",
               "Maximum Depth","Elevation","Dissolved Oxygen","pH","Conductivity",
               "Acid Neutralizing Capacity","Turbidity","Total Organic Carbon","Dissolved Organic Carbon",
               "Ammonium","Nitrate/Nitrite", "Total Nitrogen","Total Phosphorus","Chloride","Nitrate",
               "Sulfate","Calcium","Magnesium","Sodium","Potassium","Color","Silica","Hydrogen Ions",
               "Hydroxide","Calculate Ammonium","Cation Sum","Anion Sum","Anion Deficit","Base Cation Sum",
               "Ion Balance","Estimate Organic Anions","Calculated Conductivity",
               "D-H-O Calculated Conductivity","Mean Profile Water Temperature","Growing Degree Days",
               "Maximum Lake Length","Maximum Lake Width","Mean Lake Width","Fetch from North", 
               "Fetch form Northeast","Fetch from East", "Fetch from Southeast",
               "Estimated Maximum Lake Depth","Estimated Lake Volume","Estimated Mean Lake Depth",
               "Nitrogen:Phophorus Ratio")
wq<-!predictors_all%in%predictors_gis
type<-vector("character",length(predictors_all))
type[wq]<-"Water Quality"
type[!wq]<-"GIS"
data_def<-data.frame(Variables=predictors_all,Description=translation,Type=type)
# This is an unbelievably silly table... It is life sized!
kable(data_def)
#dfToImage(data_def,style="css/texttable.css",file="figure/data_def_table.jpg")
```

```{r results='asis', echo=FALSE}
state<-map_data('state')
lakes_alb<-data.frame(hkm2014Data[["AlbersX"]],hkm2014Data[["AlbersY"]])
p4s<-"+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
ll<-"+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
lakes_alb_sp<-SpatialPoints(coordinates(lakes_alb),proj4string=CRS(p4s))
lakes_dd<-spTransform(lakes_alb_sp,CRS=CRS(ll))
lakes_dd<-data.frame(coordinates(lakes_dd))
mycolor<-wes.palette(5,"Zissou2")
mycolor<-c(mycolor[1],mycolor[2],mycolor[4])
names(lakes_dd)<-c("long","lat",c())
nlaMap(state,lakes_dd,mycolor)
#ggsave("figure/nla_map.jpg",nlaMap(state,lakes_dd,mycolor),
#       width=18,units="in",dpi=600)
```
**Predicting Trophic State with Random Forests**

Random forest is a machine learning algorithm that aggregates numerous decision trees in order to obtain a consensus prediction of the response categories [@breiman2001random].  Bootstrapped sample data is recursively partitioned according to a given random subset of predictor variables and completely grown without pruning.   With each new tree, both the sample data and predictor variable subset is randomly selected.  

While random forests are able to handle numerous correlated variables without a decrease in prediction accuracy, unusually large numbers of related variables can reduce accuracy and increase the chances of over-fitting the model.  This is a problem often faced in gene selection and in that field, a variable selection method based on random forest has been succesfully applied [@diaz2006gene].  We use varselRF in R to initially examine the importance of the water quality and GIS derived variables and select a subset, the reduced model, to then pass to random forest[@diaz-uriarte2010varSelRF]. 

Using R’s randomForest package, we pass the reduced models selected with varSelRF and calculate confusion matrices, overall accuracy and kappa coeffecient [@liaw2002randomForest]. From the reduced model random forests we collect a consensus prediction and calculate a confusion matrix and summary stats.

**Model Details**

Using a combination of the `varSelRF` and `randomForest` we ran models for six combinations of variables and trophic state classifications.  These combinations included different combinations of the Chlorphyll *a* trophic states (Table 2) along with all variables and the GIS only variables (i.e. no *in situ* infromation).  The six model combinations were:

1. Chlorophyll *a* trophic state - 4 class = All variables (*in situ* water quality, lake morphometry, and landscape)
2. Chlorophyll *a* trophic state - 3 class = All variables (*in situ* water quality, lake morphometry, and landscape)
3. Chlorophyll *a* trophic state - 2 class = All variables (*in situ* water quality, lake morphometry, and landscape)
4. Chlorophyll *a* trophic state - 4 class = All variables (lake morphometry, and landscape)
5. Chlorophyll *a* trophic state - 3 class = All variables (lake morphometry, and landscape)
6. Chlorophyll *a* trophic state - 2 class = All variables (lake morphometry, and landscape)

```{r trophicStateTable, results='asis', echo=FALSE, fig.cap="Table 2. Chlorphyll a based trophic state cut-offs"}
ts_4<-c("oligo","meso","eu","hyper")
ts_3<-c("oligo","meso/eu","meso/eu","hyper")
ts_2<-c("oligo/meso","oligo/meso","eu/hyper","eu/hyper")
co<-c("<= 0.2",">2-7",">7-30",">30")
xdf<-data.frame(ts_4,ts_3,ts_2,co)
names(xdf)<-c("Trophic State (4)","Trophic State (3)","Trophic State (2)","Cut-off")
kable(xdf)
#dfToImage(xdf,style="css/texttable.css",file="figure/ts_classes_table.jpg")
```

##Results

```{r figSetup,echo=FALSE,eval=T,cache=TRUE}
#create list of all selected variables and assign a color
vars<-ls(pattern="_vars")
all_vars<-vector()
for(i in vars){
  all_vars<-c(all_vars,get(i))
}
all_vars<-unique(all_vars)

all_movies_palette<-vector()
for(i in 1:dim(namelist)[1]){
  all_movies_palette<-c(all_movies_palette,wes.palette(namelist[i,2],
                                                       namelist[i,1]))
}
all_movies_palette<-unique(all_movies_palette)
zissou<-c(wes.palette(5,"Zissou"),wes.palette(5,"Zissou2"))
col_lu<-data.frame(variables=all_vars,hexcode=sample(all_movies_palette,length(all_vars)))
```

###Model 1: 4 Trophic States ~ All Variables

```{r ALL_TS4, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts4<-iterVarSelRF(hkm2014Data[predictors_all],hkm2014Data$TS_CHLA_4,numCore=4,100,
    									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="all_ts4",time=TRUE)
```

```{r ALL_TS4_rf,echo=FALSE,eval=FALSE,cache=T}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts4_vars<-unique(unlist(all_ts4[1:100]))
all_ts4_rf<-runRandomForest(hkm2014Data[all_ts4_vars],hkm2014Data$TS_CHLA_4, ntree=10000,importance=TRUE,proximity=TRUE)
```


```{r Figure_Model1,results="asis",echo=FALSE, dev="jpeg", fig.width=8, fig.height=8}
#dfToImage(sumTable(all_ts4[1:100]),file="figure/all_ts4_table.jpg",
#         style="css/sumtable.css",width=8,include.rownames=FALSE)
#dfToImage(formatC(round(all_ts4_rf$confusion,2)),file="figure/all_ts4_confusion.jpg",
#          style="css/confusiontable.css",width=8,include.rownames=TRUE)
kable(sumTable(all_ts4[1:100]),row.names=F)
kable(formatC(round(all_ts4_rf$confusion,2)),row.names=F)


all_ts4_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%all_ts4_vars]
importancePlot(all_ts4_rf,'acc',size=10,aes(colour=all_ts4_color))
importancePlot(all_ts4_rf,'gini',size=10,aes(colour=all_ts4_color))
```

Total accuracy for Model 1 is `r round(classAgreement(all_ts4_rf$confusion[,1:4])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(all_ts4_rf$confusion[,1:4])$kappa,3)`.


###Model 2: 3 Trophic States ~ All Variables

```{r ALL_TS3, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts3<-iterVarSelRF(hkm2014Data[predictors_all],hkm2014Data$TS_CHLA_3,numCore=4,100,
  										ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="all_ts3",time=TRUE)
```


```{r all_TS3_rf,echo=FALSE,eval=FALSE,cache=T}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts3_vars<-unique(unlist(all_ts3[1:100]))
all_ts3_rf<-runRandomForest(hkm2014Data[all_ts3_vars],hkm2014Data$TS_CHLA_3, ntree=10000,importance=TRUE,proximity=TRUE)
```

```{r Figure_Model2,results="asis",echo=FALSE, dev="jpeg", fig.width=8, fig.height=8}
#dfToImage(sumTable(all_ts3[1:100]),file="figure/all_ts3_table.jpg",
#          style="css/sumtable.css",width=8,include.rownames=FALSE)
#dfToImage(formatC(round(all_ts3_rf$confusion,2)),file="figure/all_ts3_confusion.jpg",
#          style="css/confusiontable.css",width=8,include.rownames=TRUE)
kable(sumTable(all_ts3[1:100]),row.names=F)
kable(formatC(round(all_ts3_rf$confusion,2)),row.names=F)

all_ts3_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%all_ts3_vars]
importancePlot(all_ts3_rf,'acc',size=10,aes(colour=all_ts3_color))
importancePlot(all_ts3_rf,'gini',size=10,aes(colour=all_ts3_color))
```


Total accuracy for Model 2 is `r round(classAgreement(all_ts3_rf$confusion[,1:3])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(all_ts3_rf$confusion[,1:3])$kappa,3)`.

###Model 3: 2 Trophic States ~ All Variables

```{r ALL_TS2, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts2<-iterVarSelRF(hkm2014Data[predictors_all],hkm2014Data$TS_CHLA_2,numCore=4,100,
  										ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="all_ts2",time=TRUE)
```


```{r all_TS2_rf,echo=FALSE,eval=FALSE,cache=T}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
all_ts2_vars<-unique(unlist(all_ts2[1:100]))
all_ts2_rf<-runRandomForest(hkm2014Data[all_ts2_vars],hkm2014Data$TS_CHLA_2, ntree=10000,importance=TRUE,proximity=TRUE)
```

```{r Figure_Model3,results="asis",echo=FALSE, dev="jpeg", fig.width=8, fig.height=8}
#dfToImage(sumTable(all_ts2[1:100]),file="figure/all_ts2_table.jpg",
#          style="css/sumtable.css",width=8,include.rownames=FALSE)
#dfToImage(formatC(round(all_ts2_rf$confusion,2)),file="figure/all_ts2_confusion.jpg",
#          style="css/confusiontable.css",width=8,include.rownames=TRUE)
kable(sumTable(all_ts2[1:100]),row.names=F)
kable(formatC(round(all_ts2_rf$confusion,2)),row.names=F)

all_ts2_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%all_ts2_vars]
importancePlot(all_ts2_rf,'acc',size=10,aes(colour=all_ts2_color))
importancePlot(all_ts2_rf,'gini',size=10,aes(colour=all_ts2_color))
```

Total accuracy for Model 3 is `r round(classAgreement(all_ts2_rf$confusion[,1:2])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(all_ts2_rf$confusion[,1:2])$kappa,3)`.

###Model 4: 4 Trophic States ~ GIS Only Variables

```{r GIS_TS4, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts4<-iterVarSelRF(hkm2014Data[predictors_gis],hkm2014Data$TS_CHLA_4,numCore=4,100,
      								ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="gis_ts4",time=TRUE)
```

```{r GIS_TS4_rf,echo=FALSE,eval=FALSE,cache=T}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts4_vars<-unique(unlist(gis_ts4[1:100]))
gis_ts4_rf<-runRandomForest(hkm2014Data[gis_ts4_vars],hkm2014Data$TS_CHLA_4, ntree=10000,importance=TRUE,proximity=TRUE)
```

```{r Figure_Model4,results="asis",echo=FALSE, dev="jpeg", fig.width=8, fig.height=8,fig.cap="Figure 1"}
#dfToImage(sumTable(gis_ts4[1:100]),file="figure/gis_ts4_table.jpg",
#          style="css/sumtable.css",width=8,include.rownames=FALSE)

#dfToImage(formatC(round(gis_ts4_rf$confusion,2)),file="figure/gis_ts4_confusion.jpg",
#          style="css/confusiontable.css",width=8,include.rownames=TRUE)
kable(sumTable(gis_ts4[1:100]),row.names=F)
kable(formatC(round(gis_ts4_rf$confusion,2)),row.names=F)

gis_ts4_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%gis_ts4_vars]
importancePlot(gis_ts4_rf,'acc',size=10,aes(colour=gis_ts4_color))
importancePlot(gis_ts4_rf,'gini',size=10,aes(colour=gis_ts4_color))
```

Total accuracy for Model 4 is `r round(classAgreement(gis_ts4_rf$confusion[,1:4])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(gis_ts4_rf$confusion[,1:4])$kappa,3)`.

###Model 5: 3 Trophic States ~ GIS Only Variables

```{r GIS_TS3, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts3<-iterVarSelRF(hkm2014Data[predictors_gis],hkm2014Data$TS_CHLA_3,numCore=4,100,
    									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="gis_ts3",time=TRUE)
```

```{r GIS_TS3_rf,echo=FALSE,eval=FALSE,cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts3_vars<-unique(unlist(gis_ts3[1:100]))
gis_ts3_rf<-runRandomForest(hkm2014Data[gis_ts3_vars],hkm2014Data$TS_CHLA_3, ntree=10000,importance=TRUE,proximity=TRUE)
```

```{r Figure_Model5,results="asis",echo=FALSE, dev="jpeg", fig.width=8, fig.height=8}
#dfToImage(sumTable(gis_ts3[1:100]),file="figure/gis_ts3_table.jpg",
#          style="css/sumtable.css",width=8,include.rownames=FALSE)
#dfToImage(formatC(round(gis_ts3_rf$confusion,2)),file="figure/gis_ts3_confusion.jpg",
#          style="css/confusiontable.css",width=8,include.rownames=TRUE)
kable(sumTable(gis_ts3[1:100]),row.names=F)
kable(formatC(round(gis_ts3_rf$confusion,2)),row.names=F)

gis_ts3_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%gis_ts3_vars]
importancePlot(gis_ts3_rf,'acc',size=10,aes(colour=gis_ts3_color))
importancePlot(gis_ts3_rf,'gini',size=10,aes(colour=gis_ts3_color))
```

Total accuracy for Model 5 is `r round(classAgreement(gis_ts3_rf$confusion[,1:3])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(gis_ts3_rf$confusion[,1:3])$kappa,3)`.

###Model 6: 2 Trophic States ~ GIS Only Variables

```{r GIS_TS2, echo=FALSE, eval=FALSE, cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts2<-iterVarSelRF(hkm2014Data[predictors_gis],hkm2014Data$TS_CHLA_2,numCore=4,100,
  										ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
											vars.drop.num=1,outStr="gis_ts2",time=TRUE)
```

```{r GIS_TS2_rf,echo=FALSE,eval=FALSE,cache=TRUE}
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
gis_ts2_vars<-unique(unlist(gis_ts2[1:100]))
gis_ts2_rf<-runRandomForest(hkm2014Data[gis_ts2_vars],hkm2014Data$TS_CHLA_2, ntree=10000,importance=TRUE,proximity=TRUE)
```

```{r Figure_Model6,results="asis",echo=FALSE, dev="jpeg", fig.width=8, fig.height=8}
#dfToImage(sumTable(gis_ts2[1:100]),file="figure/gis_ts2_table.jpg",
#          style="css/sumtable.css",width=8,include.rownames=FALSE)
#dfToImage(formatC(round(gis_ts2_rf$confusion,2)),file="figure/gis_ts2_confusion.jpg",
#          style="css/confusiontable.css",width=8,include.rownames=TRUE)
kable(sumTable(gis_ts2[1:100]),row.names=F)
kable(formatC(round(gis_ts2_rf$confusion,2)),row.names=F)

gis_ts2_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%gis_ts2_vars]
importancePlot(gis_ts2_rf,'acc',size=10,aes(colour=gis_ts2_color))
importancePlot(gis_ts2_rf,'gini',size=10,aes(colour=gis_ts2_color))
```

Total accuracy for Model 6 `r round(classAgreement(gis_ts2_rf$confusion[,1:2])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(gis_ts2_rf$confusion[,1:2])$kappa,3)`.

###Associating Trophic State and Cyanobacteria

```{r ts_4_biov, echo=FALSE, dev="jpeg", fig.width=8, fig.height=8}
plotCdf(hkm2014Data$TS_CHLA_4,hkm2014Data$sumBioV+1,
                                     cdf_colors=zissou[6:9],y='Percent',
                                     x='Log10(Cyanobacteria Biovolume)',
                                     title=expression(paste('CDF for Chlorophyll ', 
                                     italic("a"),' Trophic States (4 Categories)')),
                                     color="Trophic State\nCategories")
```

```{r ts_3_biov, echo=FALSE, dev="jpeg", fig.width=8, fig.height=8}
plotCdf(hkm2014Data$TS_CHLA_3,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:8],
       y='Percent',x='Log10(Cyanobacteria Biovolume)',
       title=expression(paste('CDF for Chlorophyll ', italic("a"),' Trophic States (3 Categories)')),
       color="Trophic State\nCategories")
```

```{r ts_2_biov, echo=FALSE, dev="jpeg", fig.width=8, fig.height=8}
plotCdf(hkm2014Data$TS_CHLA_2,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:7],
       y='Percent',x='Log10(Cyanobacteria Biovolume)',
       title=expression(paste('CDF for Chlorophyll ', italic("a"),' Trophic States (2 Categories)')),
       color="Trophic State\nCategories")
```

```{r scatterplot,echo=FALSE, dev="jpeg", fig.width=8, fig.height=8}
scp_df<-data.frame(chla=hkm2014Data[["CHLA"]],biovp1=hkm2014Data[["sumBioV"]]+1)
scatterPlot(scp_df,xvar="chla",yvar="biovp1",zissou[8],zissou[7],zissou[6],
             title=expression(paste("Chlorophyll ", italic("a")," and Cyanobacteria Relationship")),
             x=expression(paste('Log10(Chl ', italic("a"),')')),
             y="Log10(Cyanobaterial Biovolumes + 1)")
```

#References